---
title: "Final Project"
author: "Jiner Zheng"
date: "12/9/2019"
output:
  html_document: default
  word_document: default
---

### Data Management and Preparation
```{r,message=FALSE,warning=FALSE}
library(dplyr)
library(magrittr)
library(qwraps2)
library(haven)
library(ggplot2)
library(tableone)
library(Matching)
MGM_AER_P_P <- read_dta("~/Documents/R Projects/qac 307/MGM_AER_P-P.dta")
peer_tutoring <- MGM_AER_P_P[,c(1:17,97:100)]
peer_tutoring$dfgpa <- (peer_tutoring$fgrdnum - peer_tutoring$ptermcumgpa)/peer_tutoring$ptermcumgpa 

our_summary1 <-
  list("Hours Tutored" =
         list("min" = ~ min(.data$tuthours),
              "max" = ~ max(.data$tuthours),
              "mean (sd)" = ~ qwraps2::mean_sd(.data$tuthours)),
       "Term Credit Hours Earned" =
         list("min" = ~ min(.data$termcred),
              "median" = ~ median(.data$termcred),
              "max" = ~ max(.data$termcred),
              "mean (sd)" = ~ qwraps2::mean_sd(.data$termcred)),
       "Final Grade on 4.0 scale" =
         list("min" = ~ min(.data$fgrdnum),
              "median" = ~ median(.data$fgrdnum),
              "max" = ~ max(.data$fgrdnum),
              "mean (sd)" = ~ qwraps2::mean_sd(.data$fgrdnum)),
       "Previous Term Cumulative GPA" =
         list("min" = ~ min(.data$ptermcumgpa),
              "max" = ~ max(.data$ptermcumgpa),
              "median" = ~ median(.data$ptermcumgpa),
              "mean (sd)" = ~qwraps2::mean_sd(.data$ptermcumgpa)),
       "Change in GPA compared to Previous Term" =
         list("min" = ~min(.data$dfgpa),
              "max" = ~max(.data$dfgpa),
              "median" = ~median(.data$dfgpa),
              "mean (sd)" = ~qwraps2::mean_sd(.data$dfgpa)),
       "SAT math" =
         list("min" = ~ min(.data$admsatm),
              "max" = ~ max(.data$admsatm),
              "mean (sd)" = ~ qwraps2::mean_sd(.data$admsatm)),
       "High School Rank Index" =
         list("min" = ~ min(.data$admri),
              "max" = ~ max(.data$admri),
              "mean (sd)" = ~ qwraps2::mean_sd(.data$admri)))

demo_summary1 <-
   list("Demographics" = 
          list(
            "Varisity Athlete" = ~ qwraps2::n_perc0(.data$athpart == 1),
            "Female" = ~ qwraps2::n_perc0(.data$demgender == 0),
            "Male"  = ~ qwraps2::n_perc0(.data$demgender == 1),
            "African American" = ~ qwraps2::n_perc0(.data$demafam == 1),
            "Hispanic" = ~qwraps2::n_perc0(.data$demhis == 1),
            "Greek Participant" = ~qwraps2::n_perc0(.data$demgrkpart == 1),
            "Freshman" = ~qwraps2::n_perc0(.data$acafr == 1),
            "Sophoremore" = ~qwraps2::n_perc0(.data$acaso == 1)),
        "Courses Characteristics"=
          list(
            "Recitation Course" = ~qwraps2::n_perc0(.data$acarec == 1),
            "Lab Course"=~qwraps2::n_perc0(.data$acalab == 1),
            "2004-05"=~qwraps2::n_perc0(.data$y1 == 1),
            "2005-06"=~qwraps2::n_perc0(.data$y2 == 1)))

orig_opt <- options()$qwraps2_markup
options(qwraps2_markup = "markdown")
peer_tutoring$tutpart_factor <- peer_tutoring$tutpart
whole <- summary_table(peer_tutoring, our_summary1)
by_tutor <- summary_table(dplyr::group_by(peer_tutoring, tutpart_factor), our_summary1)
bind <- cbind(whole, by_tutor)
whole2 <- summary_table(peer_tutoring, demo_summary1)
by_tutor2 <- summary_table(dplyr::group_by(peer_tutoring,tutpart_factor), demo_summary1)
bind2 <- cbind(whole2,by_tutor2)
```

##### regression model (just a glimpse)

```{r}
mod <- lm(dfgpa~tutpart, data = peer_tutoring)
summary(mod)
```


### Summary statistics 

##### Academic Performance
```{r}
print(bind,
      rtitle = "Summary Statistics of Academic Performance",
      cnames = c("Total", "Not Tutored", "Tutored"))
```

##### Demographics / Courses Characteristics
```{r}
print(bind2,
      rtitle = "Summary Statistics of Demographics and Courses Characteristics",
      cnames = c("Total", "Not Tutored", "Tutored"))
```



### General Visualizations

#### Univariate graph of percentage change in final grade

```{r}
peer_tutoring %>%
  ggplot(aes(x=dfgpa))+
  geom_density(color="maroon3")+
  theme_minimal()+
  xlab("Percentage change in final grade")+
  ggtitle("Distribution of percentage change in final grade")
```

#### Bivariate graph of final grade based on tutoring


Bivariate graph of Percentage Change in final grade 

```{r}
peer_tutoring %>%
  ggplot(aes(y=dfgpa, x=as.factor(tutpart), fill=as.factor(tutpart)))+
  geom_boxplot(alpha=0.8)+
  theme_minimal()+
  ylab("Proportion of Change in final grade based on Previous Term GPA ")+
  xlab("Tutoring Participant")+
  scale_fill_manual("Tutoring Participant",
                     labels=c("Not Tutored","Tutored"),
                     values=c("mediumpurple3","palevioletred2"))
```

##### boxplot of final grade based on tutoring

```{r}
peer_tutoring %>%
  ggplot(aes(y=fgrdnum, x=as.factor(tutpart), fill=as.factor(tutpart)))+
  geom_boxplot(alpha=0.8)+
  theme_minimal()+
  ylab("Final grade on 4.0 scale")+
  xlab("Tutoring Participant")+
  scale_fill_manual("Tutoring Participant",
                     labels=c("Tutored","Not Tutored"),
                     values=c("mediumpurple3","palevioletred2"))+
  ggtitle("Distribution of final grade based on tutoring")
```

##### bivariate graph of final grade with tutoring hours within participants in the tutoring program

```{r}
tutored <- peer_tutoring[peer_tutoring$tutpart == 1,]
tutored %>%
  ggplot(aes(x=tuthours,y=fgrdnum))+
  geom_point(color="cornflowerblue", 
             size = 2, 
             alpha = .6) +
  geom_smooth(method = "lm",
              size = 1.5,
              color = "slateblue1")+
  xlab("The number of hours being tutored")+
  ylab("Final Grade on 4.0 Scale")+
  theme_minimal()
```




##### Mirror density plot representing the distribution of previous term culumative gpa (on the bottom) and the distribution of final grade 

```{r}
ggplot(peer_tutoring,aes(x=x)) +
  # Top
  geom_density( aes(x = fgrdnum,y=..density..), fill="#69b3a2" ) +
  geom_label( aes(x=0.5, y=0.25, label="Final Grade"), color="#69b3a2") +
  # Bottom
  geom_density( aes(x = ptermcumgpa,y=-..density..), fill= "#404080") +
  geom_label( aes(x=0.5, y=-0.25, label="Previous GPA"), color="#404080") +
  theme_minimal() +
  xlab("Grade on a 4.0 Scale")
```


### Visualizations based on treatment.

##### Previous Term GPA
```{r}
peer_tutoring %>%
  ggplot(aes(x=ptermcumgpa, group=as.factor(tutpart), color=as.factor(tutpart)))+
  geom_density()+
  theme_minimal()+
  xlab("Previous Term Cumulative GPA")+
  scale_color_manual("Tutoring Participant",
                     labels=c("Tutored","Not Tutored"),
                     values=c("mediumpurple3","palevioletred2"))+
  ggtitle("")
```

##### SAT-math
```{r}
peer_tutoring %>%
  ggplot(aes(x=admsatm, group=as.factor(tutpart), color=as.factor(tutpart)))+
  geom_density()+
  theme_minimal()+
  xlab("SAT-math")+
  scale_color_manual("Tutoring Participant",
                     labels=c("Tutored","Not Tutored"),
                     values=c("mediumpurple3","palevioletred2"))+
  ggtitle("")
```

##### High school rank index
```{r}
peer_tutoring %>%
  ggplot(aes(x=admri, group=as.factor(tutpart), color=as.factor(tutpart)))+
  geom_density()+
  theme_minimal()+
  xlab("High School Rank Index")+
  scale_color_manual("Tutoring Participant",
                     labels=c("Tutored","Not Tutored"),
                     values=c("mediumpurple3","palevioletred2"))+
  ggtitle("")
```

##### Varsity Athlete
```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=athpart),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion of Varsity Athletes within each tutored group")+
  theme_minimal()
```

##### African-American
```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=demafam),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion African American within each tutored group")+
  theme_minimal()
```

##### Hispanic
```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=demhis),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Hispanic within each tutored group")+
  theme_minimal()
```

##### Male

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=demgender),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Male within each tutored group")+
  theme_minimal()
```

##### Greek Participant

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=demgrkpart),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Greek Participant within each tutored group")+
  theme_minimal()
```

##### Freshman

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=acafr),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Freshman within each tutored group")+
  theme_minimal()
```

##### Sophomore 

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=acaso),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Sophoremore within each tutored group")+
  theme_minimal()
```

##### Recitation Courses

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=acarec),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Enrolled in Recitation Courses within each tutored group")+
  theme_minimal()
```

##### Lab courses

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=acalab),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Enrolled in lab courses within each tutored group")+
  theme_minimal()
```

##### 2004-05

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=y1),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Enrolled in courses in 2004-06 within each tutored group")+
  theme_minimal()
```

##### 2005-06

```{r}
ggplot(data=peer_tutoring)+
  stat_summary(aes(x=factor(tutpart), y=y2),
               fun.y="mean", 
               geom="bar", 
               fill="purple3", 
               alpha=0.5, width=0.5)+
  xlab("Tutoring Participant")+
  ylab("Proportion Enrolled in courses in 2005-06 within each tutored group")+
  theme_minimal()
```


### Perform Matching
```{r}
mydata <- 
  peer_tutoring %>%
  dplyr::select(-c("tuthourssq","tuthourscu"))

psmodel <- glm(tutpart~admsatm+admri+termcred+athpart+demafam+demhis+demgender+demgrkpart+acafr+acaso+ptermcumgpa+acalab+acarec+y1+y2,family=binomial(), data=mydata)

summary(psmodel)
```

##### Propensity Scores before matching
```{r}
pscore <- psmodel$fitted.values

ggplot(mydata)+
geom_density(aes(x=pscore, color=factor(tutpart)))+
  xlab("Propensity Score")+
  scale_color_manual("Tutoring Participant",
                     labels=c("Not Tutored","Tutored"),
                     values=c("mediumpurple3","palevioletred2"))+
  ggtitle("Pre-Matched Propensity Scores by Tutoring")+
  theme_minimal()
```

##### Table before matching

```{r}
xvars <- c("admsatm","admri","termcred","athpart","demafam","demhis","demgender","demgrkpart","acafr","acaso","ptermcumgpa","acarec","acalab","y1","y2")
#look at a table 1
table1<- CreateTableOne(vars=xvars,strata="tutpart", data=mydata, test=FALSE)
## include standardized mean difference (SMD)
print(table1,smd=TRUE)
```

##### Perform greedy matching on propensity score

```{r}
#do greedy matching on propensity score using Match with a caliper
psmatch<-Match(Tr=mydata$tutpart,M=1,X=pscore,replace=FALSE,caliper=.2*sd(pscore))
matched<-mydata[unlist(psmatch[c("index.treated","index.control")]), ]
#get standardized differences
matchedtab1<-CreateTableOne(vars=xvars, strata ="tutpart",
data=matched, test = FALSE)
print(matchedtab1, smd = TRUE)
```
##### Propensity Scores after matching
```{r}
psmodel2 <- glm(tutpart~admsatm+admri+termcred+athpart+demafam+demhis+demgender+demgrkpart+acafr+acaso+ptermcumgpa+acalab+acarec+y1+y2,family=binomial(), data=matched)
pscore2 <- psmodel2$fitted.values

ggplot(matched)+
geom_density(aes(x=pscore2, color=factor(tutpart)))+
  xlab("Propensity Score")+
  scale_color_manual("Tutoring Participant",
                     labels=c("Not Tutored","Tutored"),
                     values=c("mediumpurple3","palevioletred2"))+
  ggtitle("After-Matched Propensity Scores by Tutoring")+
  theme_minimal()
```


##### Paired wise t-test

```{r}
y_trt<-matched$dfgpa[matched$tutpart==1]
y_con<-matched$dfgpa[matched$tutpart==0]
t.test(y_trt, y_con, paired=TRUE)

```



### Sub-analysis examining the asscoiation between tutored hours and changes in final grade within participants who were tutored

```{r,message=FALSE, warning=FALSE}
tutored <- 
  peer_tutoring %>%
  filter(tutpart == 1)

#before controlling for other covariates
model2 <- lm(dfgpa~tuthours,data = tutored)

#after controlling for covariates
model3 <- lm(dfgpa~tuthours+admsatm+admri+termcred+athpart+demafam+demhis+demgender+demgrkpart+acafr+acaso+ptermcumgpa+acalab+acarec, data = tutored)
summary(model2)
summary(model3)
```

